---
title: "Coding Training 2: Structures and Subsetting"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(gradethis)
tutorial_options(
  # use gradethis for checking
  exercise.checker = gradethis::grade_learnr
  )
knitr::opts_chunk$set(echo = FALSE)
```

## Use notes

Please make sure you run all code blocks you come across. This will help you learn and it is tracked as part of your completion when you submit your completion hash.

## Code Annotation

As you learn about R code you will encounter code chunks that you don't fully understand at first. A good way to approach this scenario is read through the code and going through the following annotation steps.

  
1. Do an initial reading of each code chunk. Note any questions that pop into your head. If you do this inline, use a hashtag (#) to add a comment.
    
2. Identify any functions you do not understand and look them up using the help function `help(function_name)` or the "Help" tab in the bottom right pane. Make sure you understand the input paramters and the form of the output. 
    
3. Identify any overall structure. Is a subtask repeated multiple times? How many times?
      
4. For each object, predict their size. Check your intuition using `dim` or `length` (explained below). You may also want to use `class` to identify the type of an object (explained below).
    
5. Write a one to two sentence description of what the code chunk does that someone without a coding background could understand.

You may want to use this framework to go through the larger code chunks in this tutorial.


## Data Management

### Vectors

The `c()` function *concatenates* items (of homogenous type) into a vector; for example, `myvector = c(1, 3, 6, 9)` creates a vector that has four elements. To learn about how many elements in the vector there are, you can use the function `length`. Read the help file for the function for more guidance.

```{r myvec, exercise = TRUE}
myvector <- c(1, 3, 6, 9)
myvector

length(myvector)
```

To access a particular element, you can use square brackets.

```{r myvec2, exercise = TRUE}
myvector <- c(1, 3, 6, 9)

myvector[2]
myvector[c(2,3)]
```

Use `c()` to create a vector called `myvector` that contains all the perfect squares (1, 4, 9, 16, $\dots$) less than or equal to 100.

```{r makeVec, exercise = TRUE}

```

```{r makeVec-solution}
myvector <- c(1, 4, 9, 16, 25, 36, 49, 64, 81, 100)
```

```{r makeVec-check}
# check code
gradethis::grade_code()
```

### Matrices

What are the dimensions of the following matrix? Use `dim` to find out. Read the help file for the function for more guidance on what this function returns if it isn't clear from the following example.

```{r makeMatrix, exercise = TRUE}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

dim(___)
```

```{r makeMatrix-solution}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

dim(mymatrix)
```

```{r makeMatrix-check}
# check code
gradethis::grade_code()
```


We can access specific rows of a matrix with `[,]` - for example, `mymatrix[2,]` will present the second row of the matrix, and `mymatrix[c(2,3),]` will yield a new matrix with the second and third rows of the original only.


```{r makeMatrix2, exercise = TRUE}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

mymatrix[c(2,3),]
```

Assign the first row of `mymatrix` to a variable named `row1`. Use the function `summary()` to  find the minimum and maximum of this row. Consult the help files for `summary` if necessary.

```{r matrixWork, exercise = TRUE}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

___ <- mymatrix[___]
summary(___)
```

```{r matrixWork-solution}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

row1 <- mymatrix[1,]
summary(row1)
```

```{r matrixWork-check}
# check code
gradethis::grade_code()
```

To get a single element in a matrix, we use row and column numbers, such as `mymatrix[r,c]`. This will also extract a submatrix if multiple entries are used for each.

```{r matrixWork2, exercise = TRUE}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

mymatrix[1,2]
```

Display the submatrix that contains the first two rows and second column of `mymatrix`. 


```{r matrixWork3, exercise = TRUE}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

mymatrix[___, ___]
```

```{r matrixWork3-solution}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

mymatrix[c(1,2), 2]
```

```{r matrixWork3-check}
# check code
gradethis::grade_code()
```

## Logicals


We can also ask questions about our data/variables. The `==` sign is a True/False test of equality. e.g. `row2 == 1` returns `TRUE` for every element in row2 that is equal to 1, and `FALSE` for elements different than 1. We can do similar things with the symbols `<`, `<=`, `>`, `>=`. 

```{r log, exercise = TRUE}
row2 <- c(1, 2, 3, 1)

row2 == 1
```


You can also combine two conditions with the AND (`&`) and OR (`|`) operators.

```{r log2, exercise = TRUE}
row2 <- c(1, 2, 3, 1)
col3 <- c(1, 1, 3, 2)

row2 >= 100 & col3 == 1

row2 == 1 | col3 < 100

sum(row2 == 1)

which(row2 == 1)
```


```{r quiz}
quiz(
  question("The sum line does the following:",
    answer("counts how many elements are in row2"),
    answer("counts how many elements are not equal to 1 in row2"),
    answer("counts how many 1s there are in row2", correct = TRUE),
    answer("sums the elements in row2"),allow_retry = TRUE,
    random_answer_order = T
  ),
  question("The which line does the following:",
    answer("sums the elements in row2 that are equal to 1"),
    answer("counts how many elements are equal to 1 in row2"),
    answer("tells the positions of the elements in row2 that are equal to 1", correct = TRUE),
    answer("tells the positions of the elements in row2 that are not equal to 1"),allow_retry = TRUE,
    random_answer_order = T
  )
)
```


We can use these conditions to select subsets.

Return the values in `col1` where the corresponding values of `col3` equal 1.

```{r subset1, exercise = TRUE}
col1 <- c(4, 5, 6, 4)
col3 <- c(1, 1, 3, 2)

col1[___]
```

```{r subset1-solution}
col1 <- c(4, 5, 6, 4)
col3 <- c(1, 1, 3, 2)

col1[col3 == 1]
```

```{r subset1-check}
# check code
gradethis::grade_code()
```


We can analyze these subsets as well. Compute the average of only the `col3` values for whom the corresponding `col1` values are 4.

```{r subset2, exercise = TRUE}
col1 <- c(4, 5, 6, 4)
col3 <- c(1, 1, 3, 2)

mean(___)
```

```{r subset2-solution}
col1 <- c(4, 5, 6, 4)
col3 <- c(1, 1, 3, 2)

mean(col3[col1 == 4])
```

```{r subset2-check}
# check code
gradethis::grade_code()
```



Find the minimum of the values in col1 whose col3 values are less than 15.

```{r subset3, exercise = TRUE}
col1 <- c(4, 5, 6, 4)
col3 <- c(10, 10, 30, 2)

min(___)
```

```{r subset3-solution}
col1 <- c(4, 5, 6, 4)
col3 <- c(10, 10, 30, 2)

min(col1[col3 < 15])
```

```{r subset3-check}
# check code
gradethis::grade_code()
```


*Note:* when your data values are text, i.e. letters or words, to use the == sign or any of these other logical operators, you need to put quotes around the text. For example, `col1 == "A"`.

## Distributiuons in R

We are often interested in generating a set of data.  If we just want a random sample from a set of numbers, the `sample()` function can be used.  For example if I want 5 random draws from the integers `1:100` **with replacement**, I could use:

```{r samp, exercise = TRUE}
## try running this a few times
sample(x=1:100, size=5, replace = TRUE)
```


There are times however, that we would like to generate sample data from a specific distribution (known as data simulation).  Since the data are simulated, you know where it came from and that allows us to check properties, algorithms, etc.  In other words, simulated data lets you check your work.

There are built in `R` functions for all of the distributions that we work with in class (See the Table below).

Each distribution has 4 functions associated with it: `r--`, `d--`, `p--`, and `q--`.
(For example, `rnorm( )`, `dnorm( )`, `pnorm( )`, and `qnorm( )`).  Each of these functions have arguments, such as the  parameters for the distribution, that you must specify.

* The `r--` function simulates random data from the distribution. 


```{r norm1, exercise = TRUE}
# For example, 10 observations from a normal distribution with mean 0 and stdev 1:
rnorm(n=10, mean=0, sd=1)
```

```{r norm2, exercise = TRUE}
# or 6 observations from a normal distribution with mean 3 and stdev 2:
rnorm(6, 3, 2)
```


* The `d--` function finds the density value of the number/vector you plug in (i.e. $f_X(X = x; \theta)$)

```{r norm3, exercise = TRUE}
# For example, The density value of 3.5 in a normal with mean 3 and stdev 1:
dnorm(x=3.5, mean=3, sd=1 )
```
```{r norm4, exercise = TRUE}
# or, the density value of 0.5 in a normal with mean 3 and stdev 1:
dnorm(0.5, 3, 1)
```
```{r norm5, exercise = TRUE}
#The density values of a vector in a normal with mean 0 and stdev 1:
dnorm(x=c(-3, -2, -1, 0, 1, 2, 3), mean=0, sd=1 )
```


* The \texttt{p--} function gives the cumulative distribution function (CDF); that is, it gives the probability of being at or below the number/vector you plug in for that distribution:  $P(X \leq x)$.

```{r norm6, exercise = TRUE}
# The probability of being less-than-or-equal to 3 in a normal with mean 0 and stdev 1:
pnorm(q=3, mean=0, sd=1)
```
```{r norm7, exercise = TRUE}
# The probability of being less-than-or-equal -3 in a normal with mean 0 and stdev 1:
pnorm(-3, 0, 1)
```
```{r norm8, exercise = TRUE}
# NOTE: you can get the probability of being greater than a number b:y 1 - p--()
# The probability of being >12 in a normal with mean 10 and stdev 3:
1 - pnorm(12, 10, 3)
```
```{r norm9, exercise = TRUE}
# or
pnorm(12, 10, 3, lower.tail=FALSE)
```


* The `q--` function gives the quantile function, or the number that marks the x$^{th}$ percentile of a specific distribution:  $P(X \leq \ ?)$ = given percentile.


```{r norm10, exercise = TRUE}
# The 50th percentile of a normal with mean 0 and stdev 1:
qnorm(p=0.50, mean=0, sd=1)
```
```{r norm11, exercise = TRUE}
# The 10th, 25th, 75th, and 90th percentile of a normal with mean -1 and stdev 0.5:
qnorm(c(0.10, 0.25, 0.75, 0.90), -1, 0.5)
```


The `q--` function for each distribution will be especially useful when we need to find critical values for our confidence intervals!

```{r norm12, exercise = TRUE}
#The 95% critical value for a normal with mean 0 and stdev 1 (z_alpha/2):
qnorm(.975, 0, 1)
```

**Note: Use the help documentation in `R` !**

* It will tell you the necessary arguments for each function (e.g. `?rbinom`, `help(rbinom)`). 

* And pay close attention! You already saw that the exponential distribution, for example, uses a different parameterization than we do.

Work through the following examples:

**Task** Create the code to generate a single (fair) coin flip using `sample` and the outcomes `H` and `T` (label your arguments inside the function)
```{r coinFlipSample, exercise = TRUE}

```

```{r coinFlipSample-solution}
sample(x=c('H', 'T'), size=1, prob=c(0.5, 0.5))
```

```{r coinFlipSample-check}
# check code
gradethis::grade_code()
```

**Task** Create the code to generate a single (fair) coin flip using `rbinom()` (label your arguments inside the function)
```{r coinFlip, exercise = TRUE}

```

```{r coinFlip-solution}
rbinom(n=1, size=1, prob=.5)
```

```{r coinFlip-check}
# check code
gradethis::grade_code()
```

**Task** Create the code to flip 3 coins 10 times where the result/output are the number of heads we saw each time we flipped the 3 coins.
```{r coinFlip10, exercise = TRUE}

```

```{r coinFlip10-solution}
rbinom(n=10, size=3, prob=.5)
```

```{r coinFlip10-check}
# check code
gradethis::grade_code()
```

**Task** Recall that the Poisson distribution is used for counting events in a fixed time period.  What is the probability of getting 4 phone calls in the next hour if on average you receive 6 phone calls an hour.

```{r Poisson, exercise = TRUE}

```

```{r Poisson-solution}
dpois(x=4, lambda=6)
```

```{r Poisson-check}
# check code
gradethis::grade_code()
```


**Task** Recall that the Exponential distribution is often used the model the length of life of electronic components.  What is the probability that a light bulb with an average lifetime of 200 hours burns out before 100 hours?

```{r Exp, exercise = TRUE}

```

```{r Exp-solution}
pexp(q=100, rate=1/200)
```

```{r Exp-check}
# check code
gradethis::grade_code()
```

\end{enumerate}


## Submit progress

As you have gone through this tutorial, your progress has been kept track of in something called a "hash". I can use this to make sure you have executed all of the code chunks and answered all of the questions. I will use this to give you a completion grade for this assignment. 

```{r context="server"}
learnrhash::encoder_logic(strip_output = T)
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = div(strong("Generate your hash and then copy it to submit to Gradescope."), br(), br())
)
```

[Go to Gradescope to submit your hash here](https://www.gradescope.com/courses/487999/assignments/2635506/submissions/new).


